Theorem-prover based verification of cache coherence protocols

### 2014-01-31 paper/Abstract.tex
stuff

Correct implementation of cache coherent systems of high importance in
industrial use. While significant efforts have been made in formally
verifying that the underlying protocol, this provides no formal
guarantees to the actual physically implemented hardware cache
coherence engine, which are generally derived manually and entail
nontrivial design-oriented refinements, \eg{} buffer sizing, caching
policies, multiplexing of distinct communication channels. Previously,
it was shown that that a formal model protocol description could be
almost automatically translated into a hardware description allowing
protocol proof to be reused for the implementation; however, any
hardware-oriented change to the design resulted in a new protocol and
required a complete reverification. We propose a metholoogy where the
proof is structured to be modular and shown mechnically allowing
designers to make desired changes to the implementation without
inducing further verification overhead. We show this via the concrete
example of an hierarchical MSI protocol and many interesting design variants.

>>>>

Correct implementation of cache coherent systems of high importance in
industrial use.

Model checking techniques which are typically used to verify
the protocols require both scaling down and simplifying the system rather than
modeling the full implementation.

 can be used to tackle these
issues by providing parameterized proofs. 
We use the Coq theorem prover to

We use the Coq theorem
prover to provide a modular proof of safety properties for two distinct cache coherence
protocols (MSI and MOESI). We identify a set of invariants which all these protocols obey  all of which obey common invariants. We use the example of an MSI protocol specified as a set of local atomic transitio. We also show how large portions of this
policies, multiplexing of distinct communication channels. Previously,
it was shown that that a formal model protocol description could be
almost automatically translated into a hardware description allowing
protocol proof to be reused for the implementation; however, any
hardware-oriented change to the design resulted in a new protocol and
required a complete reverification. We propose a metholoogy where the
proof is structured to be modular and shown mechnically allowing
designers to make desired changes to the implementation without
inducing further verification overhead. We show this via the concrete
example of an hierarchical MSI protocol and many interesting design variants.

### 2014-01-31 paper/System.tex
stuff

We have chosen as our concrete example is a multi-step hierarchical
MSI protocol~\cite{MSI}. Conceptually, our cache coherence system is
organized as in a tree-like structure with the root node as the main
memory, the leaves as the last local cache for individual processors
interfaces and the remaining nodes as intermediate cache structures
(see Figure~\ref{hier}). Whether a cache is a leaf or not is denoted
by the relation \leaf$(c)$.

Input and output happen at the leaves in the form of memory requests
issued by the associated processor $c$. These will be represented by
an unbounded input and output arrays and position value denoted
$c.\pos$, which represents the position of the next request to
process; it is set to $0$ initially.

In the MSI protocol, each node maintains the apermission state for
each address, denoting what operations may be performed on the
data. An address may be in the modified state (\Mo) representing that
it may be both read and modified, or in the shared state (\Sh) where
it may only be read, or in the Invalid state (\In), meaning it can be
neither modified or read. This forms a natural total order of
permissions (\In $<$ \Sh $<$ \Mo). As we will discuss later, this can
naturally extended to MESI and MOESI based protocols.

Each cache in addition to the data associated with each address $a$
(denoted by $c.\data[a]$). It also has a \emph{coherence state}
associated with the address (denoted by $c.\state[a]$) holding the
permissions on that address (\Mo, \Sh, or \In). At this level, each
node has a complete cache of the entire memory system; we will
restrict these caches to be partial as a refinement later. Initially,
all address in each cache is in the \In state, save for those in the
root cache which are in the \Mo state.

Non-leaf nodes also keep a copy of the coherence state of each of its
children in a \emph{directory}, which contains a the state of each of
its children caches for a particular address.  For a non-leaf cache
$p$, the state in its directory for child cache $c$ is denoted by
$p.\dstate[c][a]$, for address $a$.  The relation $\parent(c, p)$
denotes that $p$ is the parent of cache $c$.  This local version may
be stale, in that the child may have changed it state but the parent
has not yet updated itself. However, the parent state will always be a
conservative overapproximation of the child state.

@@

%The messages transmitted between the processor and the L1 cache
%is the same as in Figure \ref{req-resp} (except for the omission of the \proc{}
%field, which is indicated by the channel's index).
\begin{figure}
\begin{subfigure}{6.8cm}
\begin{tabular}{|lp{5.8cm}|}
\hline
\multicolumn{2}{|c|}{\Reqcp}\\
\hline
\from: & Current state of child\\
\myto: & Desired (higher) state of child\\
\addr: & Address associated with the request\\
\hline
\hline
\multicolumn{2}{|c|}{\Respcp}\\
\hline
\from: & Current state of child\\
\myto: & Downgraded state of child\\
\addr: & Address associated with the request\\
\data: & Data associated with the address, if necessary\\
\hline
\end{tabular}
\end{subfigure}
\begin{subfigure}{5.4cm}
\begin{tabular}{|lp{4.4cm}|}
\hline
\multicolumn{2}{|c|}{\Mpc}\\
\hline
\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
\addr: & Address associated with the request\\
\data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for messages between caches}
\label{format}
\end{figure}

In addition to the actual coherence states, each cache also maintains
a temporary \emph{wait state} for each address to deal with
transitioning between permission levels. The wait state consists of
two parts: a) a boolean representing whether the cache has sent a
request to its parent cache earlier and is waiting for a response
(denoted by $c.\wait[a]$ for address $a$) and b) one of the three
values \Mo, \Sh{} or \In representing the state upgrade that the cache
is waiting to make (denoted by $c.\waitS[a]$ for address
$a$). Similarly, each non-leaf cache has the corresponding wait states
for its directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting
for a response from its child $c$ for address $a$, and
$p.\dwaitS[c][a]$ denotes the state to which that cache $p$ is waiting
for $c$ to downgrade.

%% Both \wait{} and \dwait{} are \False{} initially
%% for each of the caches, and each of the addresses, because no cache is
%% waiting for any response from either its child or its parent
%% initially.

@@

\begin{subfigure}{\textwidth}
\centering
\begin{tabular}{|ll|}
\hline
\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, a)$: Child $c$ sending a response to downgrade to $x$ for address $a$ without any request from its parent}\\
\hline
Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[a] \wedge c.\wait[a] = \False$\\
\hline
Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,%$\\
%& $\;\;\;%
\myif c.\state[a] = \Mo \mythen c.\data[a] \myelse \_\rangle)$;\\
& $c.\state[a] \Leftarrow x$;\\
\hline
\hline
\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildRecvBadReq}$(c)$: Child $c$ receiving a request from its parent and has already downgraded}\\
\hline
Guard: & 
%$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge%$\\
%& $%
%q.\myto \ge c.\state[q.\addr]$\\
$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge q.\myto \ge c.\state[q.\addr]$\\
\hline
Action: & $\deq(\pc[c])$;\\
\hline
\end{tabular}
\caption{Other transitions at a non-root cache}
\label{childextra}
\end{subfigure}

>>>>

We have chosen as our concrete example is a hierarchical
MSI protocol~\cite{MSI}. As we will
discuss later, this can naturally extended to MESI and MOESI based protocols.
Conceptually, our cache coherence system is
organized as in a tree-like structure with the root node as the main
memory, the leaves as the last local cache for individual processors
interfaces and the remaining nodes as intermediate cache structures
(see Figure~\ref{hier}). Whether a cache is a leaf or not is denoted
by the relation \leaf$(c)$.  The relation $\parent(c, p)$
denotes that $p$ is the parent of cache $c$.

Input and output happen at the leaves in the form of memory requests
issued by the associated processor. These will be represented by
unbounded input and output arrays. For each leaf $c$, $c.\pos$ represents the
position of the next request to process; it is set to $0$ initially.

Each node maintains the a permission state for
each address, denoting what operations may be performed on the
data. An address may be in the modified state (\Mo) representing that
it may be both read and modified, or in the shared state (\Sh) where
it may only be read, or in the Invalid state (\In), meaning it can be
neither modified or read. This forms a natural total order of
permissions ($\In < \Sh < \Mo$). For leaf caches, the meaning of these
permission states is straightfoward, a load request can be processed only if the
leaf is at least in the \Sh{} state, and a store request can be processed only
if it is in the \Mo{} state. For the non-leaf cache, these states denote the
highest state present in the subtree rooted at this cache, denoting that some
cache in the subtree has the specified permissions (we are modeling inclusive
caches). Of course, all these invariants have to be enforced by the protocol.

Each cache in addition to the data associated with each address $a$
(denoted by $c.\data[a]$) also has a \emph{coherence state}
associated with the address (denoted by $c.\state[a]$) holding the
permissions on that address (\Mo, \Sh, or \In). If a cache does not have a
particular location, the it is equivalent to its permission being \In.
Initially, all addresses in each cache is in the \In state, save for those in
the root cache which are in the \Mo state.

Non-leaf nodes also keep a copy of the coherence state of each of its
children in a \emph{directory}, which contains a the state of each of
its children caches for a particular address.  For a non-leaf cache
$p$, the state in its directory for child cache $c$ is denoted by
$p.\dstate[c][a]$, for address $a$. This local version may
be stale, in that the child may have changed it state but the parent
has not yet updated itself. However, the parent state will always be a
conservative overapproximation of the child state.

@@

\begin{figure}
\begin{subfigure}{6.8cm}
\begin{tabular}{|lp{5.8cm}|}
\hline
\multicolumn{2}{|c|}{\Reqcp}\\
\hline
\from: & Current state of child\\
\myto: & Desired (higher) state of child\\
\addr: & Address associated with the request\\
\hline
\hline
\multicolumn{2}{|c|}{\Respcp}\\
\hline
\from: & Current state of child\\
\myto: & Downgraded state of child\\
\addr: & Address associated with the request\\
\data: & Data associated with the address, if necessary\\
\hline
\end{tabular}
\end{subfigure}
\begin{subfigure}{5.4cm}
\begin{tabular}{|lp{4.4cm}|}
\hline
\multicolumn{2}{|c|}{\Mpc}\\
\hline
\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
\addr: & Address associated with the request\\
\data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for messages between caches}
\label{format}
\end{figure}

In addition to the actual coherence states, each cache also maintains
a temporary \emph{wait state} for each address to deal with
transitioning between permission levels. The wait state consists of
two parts: a) a boolean representing whether the cache has sent a
request to its parent cache earlier and is waiting for a response
(denoted by $c.\wait[a]$ for address $a$) and b) one of the three
values \Mo, \Sh{} or \In representing the state upgrade that the cache
is waiting to make (denoted by $c.\waitS[a]$ for address
$a$). Similarly, each non-leaf cache has the corresponding wait states
for its directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting
for a response from its child $c$ for address $a$, and
$p.\dwaitS[c][a]$ denotes the state to which $p$ is waiting
for $c$ to downgrade. Both \wait{} and \dwait{} are \False{} initially
for each of the caches, and each of the addresses, because no cache is
waiting for any response from either its child or its parent
initially.

@@

\begin{subfigure}{\textwidth}
\centering
\begin{tabular}{|ll|}
\hline
\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, a)$: Child $c$ sending a response to downgrade to $x$ for address $a$ without any request from its parent}\\
\hline
Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[a] \wedge c.\wait[a] = \False$\\
\hline
Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,%$\\
%& $\;\;\;%
\myif c.\state[a] = \Mo \mythen c.\data[a] \myelse \_\rangle)$;\\
& $c.\state[a] \Leftarrow x$;\\
\hline
\hline
\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildDropReq}$(c)$: Child $c$ receiving a request from its parent and has already downgraded}\\
\hline
Guard: & 
%$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge%$\\
%& $%
%q.\myto \ge c.\state[q.\addr]$\\
$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge q.\myto \ge c.\state[q.\addr]$\\
\hline
Action: & $\deq(\pc[c])$;\\
\hline
\end{tabular}
\caption{Other transitions at a non-root cache}
\label{childextra}
\end{subfigure}
